---
title: "A simple tutorial using R/Bioconductor to analyze RNA sequencing data"
author: Chiachun Chiu
date: Oct 26, 2015
output: pdf_document
---

# Download reference genome

### Using mouse genome as an example

+ [NCBI](http://www.ncbi.nlm.nih.gov/projects/genome/assembly/grc/mouse/)
+ [Ensembl](http://asia.ensembl.org/Mus_musculus/Info/Index?redirect=no)


# Align reads to reference genome (using STAR as aligner)

### Downloading and installation of STAR

+ Download STAR from [github](https://github.com/alexdobin/STAR)
+ Requirement of STAR:
    - x86-64 compatible processors
    - 64 bit Linux/Mac OS X
    - More than 30 GB of RAM

```{r engine='bash', eval=FALSE}
git clone https://github.com/alexdobin/STAR.git
cd STAR
make STAR
# To include STAR-fucion
git submodule update --init --recursive
```

### Generating genome indices

+ The option `--runMode` directs **STAR** to run genome indices generation job.
+ The option `--genomeDir` specifies the path to the directory where the genome indices are stored. This directory has to be created before **STAR** run.
+ The option `--sjdbOverhang` specifies the length of the genomic sequence around the annotated junction to be used in constructing the splice junction database, where `ReadLength` is the length of the reads.
+ Including chromosome/scaffolds/patches is strongly recommended.
+ Using annotations:
    - **GTF**
    - **GFF**
    - **user-defined format**
+ For very small genome, the parameter `--genomeSAindexNbases` could be set according to the result: `min(14, log2(GenomeLength)/2) - 1`
+ For the genome with a large number of references (>5,000), the parameter `--genomeChrBinNbots` should be reduced (set to `min(18, log2(GenomeLength/NumberOfReferences))`) to reduce RAM consumption.

```{r engine='bash', eval=FALSE}
refGenomeInfoDir="RefGenome/Ensembl"
genomeVersion="Mus_musculus.GRCm38.82"
annotFile=${genomeVersion}.gtf
refgenomeFile=${genomeVersion}.fa
numThreads=12
ReadLength=75
ReadLength_1=`expr $ReadLength - 1`
runMode="genomeGenerate"
refgenomeParameterDir="ENSEMBL.Mus_musculus.release-82"
readDirs="Whole_Transcriptome"

if [[ ! -d "./$refgenomeParameterDir" ]]; then
  mkdir $refgenomeParameterDir
fi

STAR --runThreadN $numThreads \
    --runMode $runMode \
    --genomeDir ./$refgenomeParameterDir \
    --genomeFastaFiles ./$refGenomeInfoDir/$refgenomeFile \
    --sjdbGTFfile ./$refGenomeInfoDir/$annotFile \
    --sjdbOverhang $ReadLength_1

```

### Running mapping jobs

+ The option `--readFilesIn` specifies the names of the files containing the sequences to be mapped. **STAR** supports both **FASTA** and **FASTQ** files. If the read files are compressed, use the option `--readFilesCommand UncompressionCommand`.

```{r engine='bash', eval=FALSE}
for f in `cat $filelist`;
do
    echo "Now, STAR is processing ${f}.fastq......"
    if [[ $f =~ 'LPS' ]]; then
        exp=`echo $f | awk 'BEGIN {FS="_"} {print $1"_"$2}'`
    else
        exp=`echo $f | awk 'BEGIN {FS="_"} {print $1}'`
    fi

    STAR --genomeDir ./$refgenomeParameterDir \
         --readFilesIn ./$readDirs/$exp/${f}.fastq \
         --runThreadN 12 \
         --outFileNamePrefix Aligned/$f
done
```

### Converting SAM files into BAM files using samtools

```{r engine='bash', eval=FALSE}
for f in `cat $filelist`;
do
	samtools view -bS Aligned/${f}Aligned.out.sam -o Aligned/${f}.bam
done
```
